What is an Ancestral Recombination Graph?

### 2022-03-09 paper.tex
Correct references to figs (a), (b) etc
For clarity I have switched to lowercase subfigure letters, to avoid confusion with the node labels (which are capital letters). This may eventually need to be changed depending on the journal style.

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes of length $m$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event B, while that on $[0, v)$ is reached at event C.
The non-ancestral segment $[v, w)$ above
A contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
model. Note the complex full-sib relationship between B and C; this
could not be directly represented in a Griffiths-like EARG encoding
(see section XXX).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds).
(C) Remove nodes that are unary in all local trees.
(D) Rewrite edges to skip any unary nodes in local trees.
(E) The local trees for ARG (A).
(F) The local trees for ARG (D).
}
\end{figure}

% Q: would it help to talk about "tree vertices" here or would it
% just be more confusing? I worry that people would think that
% they are two different things then, and it's really important
% that they get that a node-is-a-node
The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. For example, Fig.~\ref{fig-simplification}E shows the
local trees along the genome for a simulated ARG
(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
and the path above node $D$. Before $D$ meets its common ancestor
over the interval $(93, 100]$ with the other samples it passes through
nodes $H, K, N, R$ and $S$, reflecting the corresponding
traversal through the ARG (see section XXX). Such unary nodes
in a local tree are not usually considered significant,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
\emph{common} ancestor, rather than focusing on specific
ancestors. For most purposes, we would consider an edge
directly joining sample $D$ to the node $T$ to be more
useful and informative. Simplification is the process of removing
these uninformative and (for most computational purposes)
inefficient unary paths.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. $K,N,O$ and $R$ in
Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can b
e replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, there can still be nodes that are unary in all
of the local trees. For example, node $H$ is one of the
parents of the recombinant node $D$ is still present in
Fig.~\ref{fig-simplification}B after diamond removal, but
it is unary in all of the local trees since it has an
in-degree of $1$.
Fig.~\ref{fig-simplification}C shows the resulting graph
after removing these nodes.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node $E$ is unary in the first tree
of Fig.~\ref{fig-simplification}E, but is either binary
or ternary in the other trees (recall this is a Wright-Fisher
simulation).

The final level of simplification is to remove \emph{all}
unary nodes from the local trees (Fig.~\ref{fig-simplification}D
and E). [And continue]

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% FIXME there should be earlier refs for SPRs, right?
operation~\citep{song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance [cite]
and is computable in O(??) time.

[Discussion of SPRs in an EARG]

The effects of recombination as we move from left-to-right along
the genome in an unsimplified GARG are easy to interpret.
Consider the leftmost tree in Fig~\ref{fig-simplification}.
In the first tree the parent of B is F, and in the second
tree its parent is E. Since both parents are present as
unary nodes in the tree, we simply change one edge in
order to effect the transition.
A more complex transition occurs when we have a regraft operation
that occurs above the current root (i.e., the current oldest
non-unary node). For example, the transition between the
second and third trees corresponds to node D switching
from the path through I to H, and we then must add branches
corresponding to passing through L and meeting at M.

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is not obvious how to produce an ARG given (say) a
collection of newick trees. If the internal nodes are
labelled, then we can simply input the edges for each
tree in turn, which can easily be ``squashed'' down
to a minimal representation. However, internal nodes
are rarely labelled in newick trees, and therefore
the process of contructing a graph is much harder.
In princple, we can contruct an EARG by reasoning
about the SPRs that must has occured between two
adjacent trees and postulating the recombination
nodes that must have occured. However, this process
is computationally expensive [How expensive? is this
what Espalier does?] Additionally, unless some extra
information is provided the time of the recombination
will not be known. Fundamentally, we must include the
unary nodes corresponding to recombination parents
in the trees if we wish to recover an ARG from them.

[ Cite \citep{deng2021distribution} somewhere here.]

>>>>

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(a)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes of length $m$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event B, while that on $[0, v)$ is reached at event C.
The non-ancestral segment $[v, w)$ above
A contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(b) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. (a) An ARG and equivalent local trees simulated
from a diploid Wright-Fisher
model. Note the complex full-sib relationship between B and C; this
could not be directly represented in a Griffiths-like EARG encoding
(see section XXX).
(b) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds).
(c) Remove nodes that are unary in all local trees.
(d) Rewrite edges to skip any unary nodes in local trees.
}
\end{figure}

% Q: would it help to talk about "tree vertices" here or would it
% just be more confusing? I worry that people would think that
% they are two different things then, and it's really important
% that they get that a node-is-a-node
The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. For example, Fig.~\ref{fig-simplification} shows a set of
four simulated gARGs and the local trees that can be constructed from
them (for simplicity, edge annotations are not shown)
Consider the right-most tree in Fig.~\ref{fig-simplification}a
and the path above node $D$. Before $D$ meets its common ancestor
over the interval $(93, 100]$ with the other samples it passes through
nodes $H, K, N, R$ and $S$, reflecting the corresponding
traversal through the ARG (see section XXX). Such unary nodes
in a local tree are not usually considered significant,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
\emph{common} ancestor, rather than focusing on specific
ancestors. For most purposes, we would consider an edge
directly joining sample $D$ to the node $T$ to be more
useful and informative. Simplification is the process of removing
these uninformative and (for most computational purposes)
inefficient unary paths.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. $K,N,O$ and $R$ in
Fig.~\ref{fig-simplification}a).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can b
e replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge.
Fig.~\ref{fig-simplification}b shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, there can still be nodes that are unary in all
of the local trees. For example, node $H$ is one of the
parents of the recombinant node $D$ is still present in
Fig.~\ref{fig-simplification}b after diamond removal, but
it is unary in all of the local trees since it has an
in-degree of $1$.
Fig.~\ref{fig-simplification}c shows the resulting graph
after removing these nodes.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node $E$ is unary in the first tree
of Fig.~\ref{fig-simplification}a, but is either binary
or ternary in the other trees (recall this is a Wright-Fisher
simulation).

The final level of simplification is to remove \emph{all}
unary nodes from the local trees (Fig.~\ref{fig-simplification}d).
[And continue]

[We should discuss the relationship between ARG nodes and SPRs
here. Cite \citep{deng2021distribution} too. Maybe we want to make
this a separate section, as it's a subtle and important point.]
